Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-22 text/discussion.qmd
discussion. ugh.

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. For examle, KARINA GIVE A COUPLE OF FINDINGS.

@@

We show that in 1980, when Geomarkets were being created, that Black people tended to be highly concentrated in the poorest Geomarket in the metropolitan area. Examples include, CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), TX19 -- City of Dallas, CA21 -- South & South Central Los Angeles, and PA5 -- Philadelphia County. In a handful of metropolitan areas (e.g., Cleveland), Black people remain conentrated in the poorest Geomarkets. In Chicago-Land,

Some metro areas follow a pattern whereby Black residents [KARINA -- ONE OR TWO SENTENCES]

Research question 2 analyzes the socioeconomic and racial composition of included versus excluded prospects under the hypothetical scenario that a student list purchase

Research question 2 analyzes the socioeconomic and racial composition of student list purchases that include all Geomarkets in a metro area. We examine which Geomarkets these prospects live in as a means of assessing the consequences of student list purchases that exclude particular Geomarkets. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets [OTHER RATIONALE FOR THIS RQ FROM BOBBY?].

KARINA -- ADD ONE OR TWO PARAGRAPHS SUMMARIZING RESULTS.

@@

Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of social mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. The relative sizes of these flows can be observed in @RN4722, which obtained parental income from federal income tax returns for every college in the U.S. The disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. @RN1297[p. 241-242] observes that educational credentials are property that "support their holders' claims for intermarriages with notable families\ldots, claims to monopolize socially and economically advantageous positions" and the claims made by educational credentials "are a setback for talent (charisma) in favor of property" that comes from "the desire for restricting the supply for these positions and their monopolization by the owners of educational certificates."

@@

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company, failing miserably, and being acquired by Nexus Capital. [ADD A SENTENCE ABOUT COLLEGE BOARD]

@@

As an owner-operated EM consultancy, Royall was the market leader in purchasing student lists on behalf of clients to help realize enrollment goals. Under Vista, EAB tranformed the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours, which can be conceptualized as an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading provider of free college search engine, which is an engine for leads. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista Eqity Partners -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. At the same time, EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform.

EAB combined all of these acquisitions into Enroll360:

@@

These dynamics of vertical and horizontal acquisitions in the enrollment management market yield policy concerns:

- START HERE TUESDAY: higher prices; funneling prospects only to colleges that are willing/able to pay for subscriptions.

to become a new kind of student list vendor.

>>>>

---
#title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
#author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. [KARINA SUMMARIZE FINDINGS] We show that in 1980, when Geomarkets were being created, that Black people tended to be highly concentrated in the poorest Geomarket in the metropolitan area. Examples include, CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), TX19 -- City of Dallas, CA21 -- South & South Central Los Angeles, and PA5 -- Philadelphia County. In a handful of metropolitan areas (e.g., Cleveland), Black people remain conentrated in the poorest Geomarkets. In Chicago-Land,  [KARINA -- ONE OR TWO SENTENCES]

@@

Research question 2 analyzes the racial and socioeconomic composition of student list purchases. We examine purchases that include all Geomarkets around a metropolitan area in order to assess the potential consequences if including/excluding particular Geomarkets from the purchase. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets. KARINA -- ADD TEXT SUMMARIZING RESULTS.

@@

Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of social mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. The relative sizes of these flows can be observed in @RN4722, which obtained parental income from federal income tax returns for every college in the U.S. The disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297[p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

@@

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company, failing miserably, and being acquired by Nexus Capital. [ADD A COUPLE SENTENCES ABOUT COLLEGE BOARD; MAYBE GEODEMOGRAPHIC SEGMENT]

@@

As an owner-operated EM consultancy, Royall was the market leader in purchasing student lists on behalf of clients to help realize enrollment goals. Under Vista, EAB tranformed the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours, which can be conceptualized as an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading provider of free college search engine, which is an engine for leads. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista Eqity Partners -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. At the same time, EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform. EAB combined all of these acquisitions into Enroll360:

@@

We argue that an under-studied and under-regulated third-party EM industry has negative consequences for colleges and for students. Accross multiple student list products, we see a growing number of filters that recommend which prospects to target based on the beheavior of previous cohorts of students. The ACT/Encoura "Enrollment Predictor" filter enables colleges to filter prospects based on the predicted probability of enrollment, which is created based on analyzing the enrollment decisions of previous cohorts of students [schmidt_2019]. Based on the Claritas/PRIZM market segmentation system that categorized zip codes into groups useful for merchants [@RN4957], The College Board Geodemographic Segment filters categorizes each Census Tract into one of about two dozen clusters based on the college-going behavior of previous cohorts [@RN4565]. A @vista_eab_ml video promoting EAB's "Pipeline Analytics" product to investors states, "what we are doing is using artificial intelligence and machine learning to help schools build list of students may be more more likely to enroll and persist in their university." The product "identifies students that would be a good match for a school based on their similarity to students who have succeeded there before and then provides them [the students] information about those institutions so that they choose to apply and ultimately enroll." Each of these products are based on the underlying logic of correlations and homophily discussed by @RN4975 and applied by @RN4982. These products identify correlations in a historic snapshot of students and apply those relationships to make recommendations about which prospects to target in the present. As such, all of these technologies amplify the effect of previous inequality in student demand, which is itself a consequence of structural inequality in educational opportunity.

Market dynamics in the EM industry may also affect competition, prices, and choice. To the extent that colleges need to buy expensive software to recruit prospects contained in an associated database, those costs will ultimately be passed on to students in the form of higher prices are reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that don't have vertical inputs -- e.g., control over proprietary pools of prospects -- cannot compete. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that are willing and able to pay for subscriptions, which raises policy concerns about student choice.

## Developing a Knowledge Infrastructure

Regulation of the higher education industry should be informed by research that investigates the "totality of relevant actors" [@RN527, p. 148] in the organizational field. The paucity of research about third-party vendos that structure college access is substantially due to difficulty obtaining data for empirical analyses.

Knowledge infrastructures are "systems of observation and measurement" [@RN5025, p. 743] that provide the foundation for scalable research. They collect, process, and distribute data in ways that "enable certain kinds of knowledge production while simultaneously channeling researchers away from questions not readily answerable within\ldots that infrastructure" [@RN5025, p. 742]. The knowledge infrastructure of education research largely consists of data about students, schools, and colleges. It has been shaped by the ascendance of economics [@RN5036], which incorporates district, state, and federal administrative data to evaluate the effects of schools and policies on student outcomes [@RN5043]. The research community mobilized to investigate for-profit colleges because for-profit colleges and students were included in ongoing data collection [@RN5022]. By contrast, researchers ignored the growth of private, for-profit, third-party vendors (e.g., PowerSchool, Parchment) because these organizations are excluded from data collections known to education researchers.

We argue that sociologists and adjacent scholars should expand the domain of the education research knowledge infrastructure to include third-party vendors in enrollment management, and edtech more broadly. This is a difficult task. Student unit record data owned by district, state, and federal agencies is simultaneously granular -- can evaluate a particular policy -- and macro -- can speak to broad trends in the education sector. By contrast, obtaining granular data about third-party products in education is difficult because these data are owned by private interests that do not want researchers to interrogate their products.

We see two broad avenues for research: case studies involving primary data collection -- including this study -- which tend to have a granular focus on particular products or vendors; and secondary data analysis, which may have a more macro focus. Case study research traditionally relies on interviews, ethnographies, and document analysis. Another approach to data collection is to issue records requests to public entities that contract with third-party entities. For example, @RN5039 analyze public university contracts with online program managers (OPMs). We issued public records requests to obtain tabular data about student list purchases. However, this process was labor intensive, requiring the pro bono efforts of several law firms. Public records requests seem better suited to obtaining contracts. Text analysis of publicly available websites and social media provides another avenue for data collection. Here, data collection is efficient but data processing is often labor intensive. 
Data collected for case study research typically does not meet the knowledge infrastructure standard of repeated, ongoing data collection that can be the foundation for scalable research. Nevertheless, case study research is important for granular analyses of particular products -- like Intersect, Enroll360,  that yield important insights about underlying mechanisms about how vendors structure college access.

A growing set of "market intelligence" data providers serve the information needs of the investment community by providing data on private firms and deals involving private firms. These databases may provide the foundation for scalable research on the role of private firms and private equity investors in education. These databases tend to include three "types" of data: (1) *firm-level data* (e.g., location, employees, financials); (2) *acquisitions*, encompassing who acquired whom and resulting ownership structure); and (3) *investments*, which include amounts/valuations/stake percentages of private equity investments. Some providers provide all three types of data (e.g., [*S&P Capital IQ*](https://www.spglobal.com/market-intelligence/en), [*LSEG Workspace*](https://www.lseg.com/en)). Other providers specialize. For example, [*PitchBook*](https://pitchbook.com/) is a leading provider of data about investments in private companies by private equity firms, while [*Orbis*](https://login.bvdinfo.com/R1/Orbis) is the leading provider of firm-level data for non-U.S. private firms [@RN5038]. Interestingly, the database subscriptions operate similar to contemporary student list products; an interactive, user-facing platform is wrapped around several sources of proprietary data. University libraries provides full or partial access to some subscriptions, while others must be purchased. These databases are often utilized for articles in management journals that focus on private equity markets [CITE], or patterns of acquisitions [CITE]. Considering the growing share of economic activity that  [HOW THESE DATABASES HAVE BEEN USED; HOW THEY MAY BE USED]
